A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-20 summary.tex
edited summary

\subsection{Other dust-drag
  instabilities}\label{dust_as_thermo}

@@

%give numerical example 
We ran our one-fluid stability code to find unstable modes  
similar to that described by \cite{loren15}. We use the same
parameters reported by \citeauthor{loren15}, as follows. 
%we set $p=q=0$, which eliminates VSI and minimizes streaming
%instabilities as there is no radial pressure gradient at the disk
%midplane.   
The total metalicity $Z=0.01$ and dust scale-height $H_\mathrm{d} =
0.4H_\mathrm{g}$, giving a maximum dust-to-gas ratio
$\mathrm{max}(\epsilon) = 0.025$. The midplane stopping time is 
$t_\mathrm{s0}=0.02\OmK^{-1}$. The disk parameters are $p=-1.5,\, q=-1$, and  
$\hgas=0.05$. We set $k = \pi/H_\mathrm{d}$.

@@

\section{Summary}\label{summary}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. 
In the limit of perfect dust-gas coupling, the dust content of a fluid 
parcel $\rhod/\rhog$ is conserved. This is analogous to entropy being 
conserved following a adiabatic, pure gas. %s = s(\rhod/\rhog)
We explicitly show that for 
a locally isothermal, dusty gas that the evolutionary equation of 
$\rhod/\rhog$ may be recast into an effective energy equation identical
in form to that in standard adiabatic hydrodynamics. We find the 
natural definition of the entropy of the dusty gas is 
\begin{align*}
  s  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
This allows us to define the buoyancy frequency of an isothermal 
dusty gas.

%
%finite coupling/heat source

Finite dust-gas friction causes a relative drift between the two
components of the mixture. Then $\rhod/\rhog $ is no
longer conserved following a fluid parcel, as it is allowed to
exchange dust particles with neighboring parcels owing to a non-zero
stopping time $\tstop$. This is analogous to heat exchange between a 
pure gas parcel and its surroundings. Thus, finite dust-gas friction
may be interepreted as an energy/entropy source term for the dusty
gas.

%
In this first study, we apply this equivalence principle to study the
axisymmetric stability of locally isothermal, protoplanetary disks
perfectly coupled to dust ($\tstop=0$). This is so that we may 
compute exact, steady equilibria for analyses. Furthermore,
in this limit the dusty problem is \emph{identical} to adiabatic, pure
gas dynamics.

We obtain the Solberg-Hoiland criteria to assess
axisymmetric stability when the equation of state is strictly
isothermal. We show that {\bf dust-settling cannot lead to
  axisymmetric instability, however large the associated vertical
  shear} \citep[cf. \emph{non-axisymmetric} Kelvin-Helmholtz instabilities
  induced by  dust-settling, ][]{lee10}. We also show that
sharp edges in the dust-to-gas ratio cannot persist in protoplanetary
disks, as these imply a sharp entropy gradient and may thus destablize
the disk.

@@

For locally isothermal disks $c^2_s=c_s^2(r)$ we generalize the vertical
shear instability (VSI) to include dust. We find that dust-loading
generally has a stabilizing effect. In our disk models
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed at heights where vertical 
buoyancy dominates over vertical shear. This confirms our 
identification of the entropy and buoyancy of an isothermal, dusty
gas.

>>>>

\subsection{New dust-drag
  instabilities}\label{dust_as_thermo}

@@

%give numerical example 
We have applied our one-fluid stability framework to find unstable modes  
similar to that described by \cite{loren15}. We use the same
parameters reported by \citeauthor{loren15}, as follows. 
%we set $p=q=0$, which eliminates VSI and minimizes streaming
%instabilities as there is no radial pressure gradient at the disk
%midplane.   
The total metalicity $Z=0.01$ and dust scale-height $H_\mathrm{d} =
0.4H_\mathrm{g}$, giving a maximum dust-to-gas ratio
$\mathrm{max}(\epsilon) = 0.025$. The midplane stopping time is 
$t_\mathrm{s0}=0.02\OmK^{-1}$. The disk parameters are $p=-1.5,\, q=-1$, and  
$\hgas=0.05$. We set $k = \pi/H_\mathrm{d}$.

@@

\section{Summary}\label{summary}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $tstop \to 0$,  
 dust is entrained in 
the gas, and $\rhod/\rhog$ is conserved following the dusty flow. 
This is analogous to entropy being conserved following a adiabatic,
pure gas.

For $\tstop\neq 0$, the dust content of a
fluid parcel of the dusty mixture no longer conserved. The parcel is
is allowed to exchange dust particles with neighboring parcels owing
to finite dust-gas friction. This is analogous to heat exchange between a 
pure gas parcel and its surroundings. %Thus, finite dust-gas friction
%may be interepreted as an energy source for the dusty
%gas.

We explicitly show that for a locally isothermal, dusty gas that the
evolutionary equation of  $\rhod/\rhog$ may be replaced by an 
effective energy equation. We find the 
natural definition of the entropy of the dusty gas is 
\begin{align*}
  s  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
%This allows us to define the buoyancy frequency of an isothermal 
%dusty gas. 
In our framework, finite dust-gas friction appears as an energy
source term.  This analogy with standard adiabatic
hydrodynamics allow us to obtain dusty analogs of gaseous
instabilities, and provide thermodynamical interpretations of  
dust-drag instabilities.

%
In this first study, we apply this equivalence principle to study the
axisymmetric stability of isothermal, protoplanetary disks
perfectly coupled to dust ($\tstop=0$). This is so that we may 
compute exact, steady equilibria for analyses. Furthermore, 
in this limit the dusty problem is \emph{identical} to adiabatic, pure
gas dynamics.

We obtain the Solberg-Hoiland criteria to assess 
axisymmetric stability when the equation of state is strictly
isothermal. We show that {\bf dust settling, by itself, cannot lead to
  axisymmetric instability, however large the associated vertical
  shear} \citep[cf. \emph{non-axisymmetric} Kelvin-Helmholtz instabilities
  in dusty disks][]{lee10}. We show that
sharp edges in the dust-to-gas ratio cannot persist in protoplanetary
disks, as these imply a sharp entropy gradient and may thus destablize
the disk.

@@

For locally isothermal disks $c^2_s=c_s^2(r)$ we generalize the vertical
shear instability (VSI) to include dust. We find that dust-loading
generally has a stabilizing effect. In our disk models
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed at heights where vertical 
buoyancy dominates over vertical shear. These results are readily
understood by comparing the vertical shear rate with the buoyancy of
an isothermal, dusty gas which we have correctly identified.

We also discuss future applications of our thermodynamic framework to 
study dusty protoplanetary disks. In particular, we emphasize  that the 
physical origin for any instability associated with strong dust-gas
drag is a phase lag between the pressure and total density evolution
of fluid elements. This leads to growing oscillations as positive work
is done, provided by dust-gas drag, which is equivalent to a `heat'
source. %We demonstrate a potentially new dust-drag instability in
%stratified disks that may explain

%We also applied
%our framework to find unstable modes in dusty, stratified disks
%similar to that reported recently by \cite{loren15}.
